Combined annotation exploration for SCPCP000006

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
fix typos

| first level annotation | second level annotation | selection of the cells                                                                        | marker genes for validation | CNV validation                             |
| ---------------------- | ----------------------- | --------------------------------------------------------------------------------------------- | --------------------------- | ------------------------------------------ |
| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr`                                   | `VWF`                       | no CNV                                     |
| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr`                                        | `PTPRC`, `CD163`, `CD68`    | no CNV                                     |
| normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr.lower`     | `CDH1`, `PODXL`, `LTL`      | no CNV                                     |
| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr.upper`            | `VIM`                       | no CNV                                     |
| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr.upper`                                         | `VIM`                       | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CITED1`                    | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CDH1`                      | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| unknown                | -                       | the rest of the cells                                                                         | -                           | -                                          |

@@

#### do_BoxPlot_mean

@@

- _normal_ cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV.
We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
Indeed, we know that false positive CNV can be observed in a cell type specific manner.
Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` > `r params$predicted.celltype.threshold` as `normal`.

@@

- _cancer_ cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV

@@

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain _unknown_

@@

- _cancer stroma_: We define as _cancer stroma_ all cancer cells from the stroma compartment.

- _blastema_,: we defined as _blastema_ every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
We know that these _mesenchymal_ cells are cells from the cap mesenchyme that are not expected to be in a mature kidney.
These blastema cells should express higher _CITED1_ and/or _NCAM1_.

- _cancer epithelium_: we defined as _cancer epithelium_ all cancer cells that are neither stroma nor blastemal cells.
We expect these cells to express epithelial markers.
Their predicted cell type should correspond to more mature kidney epithelial subunits.

@@

- identify chromosome and cell types that might be more sensitive to _FALSE_ positive CNV, such as the _chr6p_ which is a hub for immune genes

@@

Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). This represent the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).

Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis > 0, in red) or a `loss` (y-axis < 0, in blue).

@@

Comparing with the profile in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:

- _chr8q_, _chr12_, _chr13_, _chr18q_ gains

- _chr11q_, _chr16_, _chr17q_ loss

@@

- we couldn't detect any _chr4q_ loss as described in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

- we identified 2 losses in _chr5_ and _chr6_ that do not show up in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
These can be `infercnv` induced _FALSE_positive CNV.

Our major concern is the pattern of the _chr1_. 
We do see a tendency of _chr1p_ loss and _chr1q_ gain but it is not as pronounced as in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
_chr1q_ gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al, Curr Opin Pediatr (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)).
It is however important to notice that the profile shown in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol. 
So far, we do not know how the pre-operative chemotherapy impact the clonal selection of specific CNV.

@@

As described in [Nelson et al, Curr Opin Pediatr (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), 
[Phelps et al, Children (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and 
[Zheng et al, Front. Oncol (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), 
LOH at _chr1p_, _chr11p_ and _chr16q_ as well as _chr1q_ gain are prognostic factors.

@@

In our study, we validate the association of _chr1q_ gain, _chr11q_ loss and _chr14q_ loss with poorer prognosis.

@@

As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
We those check the repartition of CNV in samples that have been pre-treated (i.e `Resection post chemotherapy`) with samples that have been removed surgically before any treatment (i.e `Upfront resection`).

@@

Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, 
including a tendency of more _chr1q_ gain samples after chemotherapy. 
This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

@@

Immune cells are well annotated based on _PTPRC_ expression.

@@

Endothelial cells are well annotated based on _VWF_ expression.

@@

(*) To the best of our knowledges, there is no Wilms tumor specific marker of stroma cells.
They should express in some extend similar markers as normal stromal cells including _VIM_, _THY1_ and _COL6A3_.

@@

_VIM_ do not seem to be neither specific nor universal.

@@

_COL6A3_ expression is higher in Wilms tumor stroma cells and expressed across a wider range of samples.

@@

_THY1_ do not seem to be neither specific nor universal.

@@

(*) To the best of our knowledges, there is no Wilms tumor specific and universel (i.e. expressed by all cells in every sample) marker of blastema cells.
We expect however blastema cells to express higher levels of stemness markers such as _CITED1_, _SIX1/2_, _NCAM1_.

@@

Blastema cells are enriched in _CITED1_+ cells, but some samples do not express it at all.

@@

_NCAM1_ seems to be more cancer specific and broadly expressed among samples.

@@

As expected, epithelial cells express higher expression of _CDH1_.

@@

Confirm how many samples we have annotations for:
```{r}
length(unique(annotations_table$scpca_sample_id))
```

@@

- The mean CNV profile of the Wilms tumor cohort seems reasonable in regards to the one published in [Cresswell et al, BioRXiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
We identify some commonalities but also some divergence. 
They can be due to technical limitation of our CNV inference but can also reflect some specificities of this cohort (e.g. pre/post treatment samples).
It would be really useful to compare for some sample the `infercnv` result to a ground truth such as SNP Array (to be requested).

>>>>

| first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
|----|----|----|----|----|
| normal | endothelial | `compartment == "endothelium" & predicted.score > pred.thr` | `VWF` | no CNV |
| normal | immune | `compartment == "immune" & predicted.score > pred.thr` | `PTPRC`, `CD163`, `CD68` | no CNV |
| normal | kidney | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr.lower` | `CDH1`, `PODXL`, `LTL` | no CNV |
| normal | stroma | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr.upper` | `VIM` | no CNV |
| cancer | stroma | `compartment == "stroma" & cnv_score > cnv.thr.upper` | `VIM` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer | blastema | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper` | `CITED1` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer | epithelial | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper` | `CDH1` | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| unknown | \- | the rest of the cells | \- | \- |

@@

#### `do_BoxPlot_mean`

@@

-   *normal* cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV. We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score. Indeed, we know that false positive CNV can be observed in a cell type specific manner. Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` \> `r params$predicted.celltype.threshold` as `normal`.

@@

-   *cancer* cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV

@@

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` *cancer* cells, `r table(cell_type_df$first.level_annotation)["normal"]` *normal* cells. `r table(cell_type_df$first.level_annotation)["unknown"]` cells remain *unknown*

@@

-   *cancer stroma*: We define as *cancer stroma* all cancer cells from the stroma compartment.

-   *blastema*,: we defined as *blastema* every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`. We know that these *mesenchymal* cells are cells from the cap mesenchyme that are not expected to be in a mature kidney. These blastema cells should express higher `CITED1` and/or `NCAM1`.

-   *cancer epithelium*: we defined as *cancer epithelium* all cancer cells that are neither stroma nor blastemal cells. We expect these cells to express epithelial markers. Their predicted cell type should correspond to more mature kidney epithelial subunits.

@@

-   identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as the *chr6p* which is a hub for immune genes

@@

Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). This represent the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).

Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis \> 0, in red) or a `loss` (y-axis \< 0, in blue).

@@

Comparing with the profile in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:

-   *chr8q*, *chr12*, *chr13*, *chr18q* gains

-   *chr11q*, *chr16*, *chr17q* loss

@@

-   we couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

-   we identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). These can be `infercnv` induced \_FALSE_positive CNV.

Our major concern is the pattern of the *chr1*. We do see a tendency of *chr1p* loss and *chr1q* gain but it is not as pronounced as in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). *chr1q* gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al, Current Opinion in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)). It is however important to notice that the profile shown in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol. So far, we do not know how the pre-operative chemotherapy impact the clonal selection of specific CNV.

@@

As described in [Nelson et al, Current Opinions in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), [Phelps et al, Children (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and [Zheng et al, Frontiers in Oncology (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), LOH at *chr1p*, *chr11p* and *chr16q* as well as *chr1q* gain are prognostic factors.

@@

In our study, we validate the association of *chr1q* gain, *chr11q* loss and *chr14q* loss with poorer prognosis.

@@

As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). We those check the repartition of CNV in samples that have been pre-treated (i.e `Resection post chemotherapy`) with samples that have been removed surgically before any treatment (i.e `Upfront resection`).

@@

Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, including a tendency of more *chr1q* gain samples after chemotherapy. This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

@@

Immune cells are well annotated based on `PTPRC` expression.

@@

Endothelial cells are well annotated based on `VWF` expression.

@@

(\*) To the best of our knowledge, there is no Wilms tumor specific marker of stroma cells. They should express in some extend similar markers as normal stromal cells including *VIM*, *THY1* and *COL6A3*.

@@

*VIM* do not seem to be neither specific nor universal.

@@

*COL6A3* expression is higher in Wilms tumor stroma cells and expressed across a wider range of samples.

@@

*THY1* do not seem to be neither specific nor universal.

@@

(\*) To the best of our knowledge, there is no Wilms tumor specific and universal (i.e. expressed by all cells in every sample) marker of blastema cells. We expect however blastema cells to express higher levels of stemness markers such as `CITED1`, `SIX1/2`, `NCAM1`.

@@

Blastema cells are enriched in *CITED1*+ cells, but some samples do not express it at all.

@@

`NCAM1` seems to be more cancer specific and broadly expressed among samples.

@@

As expected, epithelial cells express higher expression of `CDH1`.

@@

Confirm how many samples we have annotations for:

```{r}
length(unique(annotations_table$scpca_sample_id))
```

@@

-   The mean CNV profile of the Wilms tumor cohort seems reasonable in regards to the one published in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). We identify some commonalities but also some divergence. They can be due to technical limitation of our CNV inference but can also reflect some specificities of this cohort (e.g. pre/post treatment samples). It would be really useful to compare for some sample the `infercnv` result to a ground truth such as SNP Array (to be requested).
